---
title: "setup for everyones report"
output: html_document
---

### SECTION 1: Libraries


# ```{r}
# #may need to do this to load some libaries
#
#
# if (!requireNamespace("BiocManager", quietly = TRUE)) {
#   install.packages("BiocManager")
# }
#
# # Install the human annotation database
# BiocManager::install("org.Hs.eg.db", ask = FALSE)
#
# # Then load the packages
# library(AnnotationDbi)
# library(org.Hs.eg.db)
#
# ```

```{r}
#loaded Libraries 

library(dplyr)
library(readr)
library(ggplot2)
library(tidyr)
library(e1071)
library(purrr)
library(survival)
library(survminer)
library(AnnotationDbi)
library(org.Hs.eg.db)
library(DESeq2)
library(pheatmap)
library(dplyr)
library(tidyr)
library(tibble)
library(purrr)
library(e1071)
```

### SECTION 2: Loading Data

```{r}
# Loading in our Expression Data
expr_raw <- read.csv(
  "RNAseq_LIHC.csv",
  header = TRUE,
  row.names = 1,          # gene IDs become rownames
  check.names = FALSE     # keep TCGA IDs untouched
)

dim(expr_raw)
expr_raw[1:5, 1:5]

# Make expression matrix numeric
expr_mat <- as.matrix(expr_raw)
mode(expr_mat) <- "numeric"

# Loading in our Clinical Data
clinical <- read_tsv(
  "data_clinical_patient.txt",
  comment = "#",
  show_col_types = FALSE
)



clinical <- as.data.frame(clinical)

rownames(clinical) <- clinical$PATIENT_ID

# Keeping only matching patient barcodes
colnames(expr_mat) <- substr(colnames(expr_mat), 1, 12)

common_ids <- intersect(colnames(expr_mat), rownames(clinical))
length(common_ids)

expr_filt    <- expr_mat[, common_ids]
clinical_filt <- clinical[common_ids, ]

# Check they are aligned
stopifnot(all(colnames(expr_filt) == rownames(clinical_filt)))

dim(expr_filt)
dim(clinical_filt)
```

### SECTION 3: Normalization

```{r}
# Building DESeqDataSet using filtered & matched data
dds <- DESeqDataSetFromMatrix(
  countData = expr_filt,      # <- use expr_filt
  colData   = clinical_filt,  # <- use clinical_filt
  design    = ~ 1             # no specific design yet
)

# Keeping only genes with 10 or more counts
keep <- rowSums(counts(dds)) >= 10
dds  <- dds[keep, ]

# Normalized counts
dds <- estimateSizeFactors(dds)
norm_counts <- counts(dds, normalized = TRUE)

# Variance stabilized transform (for downstream analysis such as clustering)
vsd <- vst(dds, blind = TRUE)
expr_vst <- assay(vsd)

dim(expr_vst)
expr_vst[1:5, 1:5]

# Should flter our hghest varance 


```


### Section 4: Create function to convert gene codes to actual gene names

```{r}
get_gene_symbols <- function(ens_ids) {
  # Remove version numbers
  ens_clean <- sub("\\..*", "", ens_ids)
  
  # Map Ensembl IDs → gene symbols
  symbol_map <- mapIds(
    org.Hs.eg.db,
    keys = ens_clean,
    keytype = "ENSEMBL",
    column = "SYMBOL",
    multiVals = "first"
  )
  
  # Replace any NA with original ENSG ID if no symbol found
  symbol_map[is.na(symbol_map)] <- ens_ids[is.na(symbol_map)]
  
  # Return just the values as a plain vector
  unname(symbol_map)
}


```

### SECTION 5: PCA ###
```{r}
expr_vst_t <- t(expr_vst)

# Perform PCA
pca_res <- prcomp(expr_vst_t, center = TRUE, scale. = FALSE)

var_explained <- pca_res$sdev^2 / sum(pca_res$sdev^2)
var_explained_percent <- round(var_explained * 100, 2)

var_explained_percent[1:5]   

# Loadings for PC1    chat generated 
pc1_loadings <- pca_res$rotation[, 1]  
pc1_abs <- abs(pc1_loadings)
top10_genes_pc1 <- names(sort(pc1_abs, decreasing = TRUE))[1:10]
top10_genes_pc1_symbols <- get_gene_symbols(top10_genes_pc1)
top10_genes_pc1_symbols

```

PCA is useful in RNA-seq expression analysis because gene expression data are high-dimensional and continuous, with thousands of genes varying across samples. PCA reduces this complexity by identifying the major axes of variation that summarize overall biological patterns, such as immune activity, tumor purity, or cell-cycle programs, and helps reveal whether clinical factors (like early vs late cancer stage) naturally separate in the data. However, this same approach does not make sense for the mutation dataset, because mutation matrices are extremely sparse and binary—most genes are unmutated in most patients, producing almost all zeros with no continuous gradients of variation. Without meaningful covariance or continuous structure, PCA cannot extract biologically relevant components from mutation data. In short, PCA works well for continuous gene expression matrices but is not appropriate for sparse 0/1 mutation profiles.



CYP3A4 is a crucial enzyme in the liver, and its expression can significantly impact liver cancer. The gene is involved in the metabolism of more than 50% of prescription drugs, and its polymorphisms can influence how patients metabolize chemotherapeutic agents, affecting treatment outcomes and the risk of adverse effects. Genetic testing for CYP3A4 variants may offer valuable insights for personalized cancer therapy, as these variants can alter enzyme activity and influence the effectiveness of chemotherapy. 

hypothesis: The genes with the most varaince will likley have an impact on cancer



use PCA1-PCAn to do analysis like clustering - then when we want to do analysis we can look at the top genes that make up each pca to see what genes supply the most variance

### SECTION 6   ###

```{r}
# Scree plot
df_scree <- data.frame(
  PC = paste0("PC", 1:length(var_explained_percent)),
  Variance = var_explained_percent
)

ggplot(df_scree[1:9, ], aes(x = PC, y = Variance)) +
  geom_bar(stat = "identity", fill = "steelblue") +
  geom_text(aes(label = Variance), vjust = -0.3) +
  theme_minimal() +
  ylab("% Variance Explained") +
  ggtitle("PCA Variance Explained")

```
  
### SECTION 7   ###



```{r}
loadings <- pca_res$rotation   

weights <- var_explained  

influence_scores <- rowSums(abs(loadings) * weights)

top_genes <- sort(influence_scores, decreasing = TRUE)[1:20]
top_gene_names <- get_gene_symbols(names(top_genes))

top_gene_names

```
AWSOME: clearly from ths out top gene DOES affect cancer so we should be able to do analyss on ths!!!

The GSTM1 gene is crucial for the detoxification of xenobiotics, carcinogens, and oxidative stress products. It is primarily located in the liver and can be present or absent, with the presence of at least one copy indicating a functional gene. The absence of GSTM1 can lead to a reduced detoxification capacity, increasing the risk of chronic diseases such as cancer, chemical sensitivity, and atopic asthma. To mitigate these risks, dietary changes such as consuming cruciferous vegetables and allium foods can be beneficial.

### SECTION 8  ###


```{r}
df_influence <- data.frame(
  Gene = get_gene_symbols(names(top_genes)),
  Score = top_genes
)

ggplot(df_influence, 
       aes(x = reorder(Gene, Score), y = Score)) +
  geom_bar(stat = "identity", fill = "darkred") +
  coord_flip() +
  theme_minimal() +
  ggtitle("Weighted Influence Scores Across All PCs")

```
### SECTION 9   ###

```{r}
top20_loadings <- loadings[names(top_genes), 1:10]
pheatmap(top20_loadings,
         cluster_cols = TRUE,
         cluster_rows = TRUE,
         main = "Loadings of Top Influential Genes (PC1–PC10)")

```



### SECTION 11   ###


```{r}
var_genes <- order(rowVars(expr_vst), decreasing = TRUE)[1:500]
cor_mat <- cor(t(expr_vst[var_genes, ]))

pheatmap(cor_mat,
         show_rownames = FALSE,
         show_colnames = FALSE,
         main = "Gene–Gene Correlation (Top 500 Most Variable Genes)")

```

### SECTION 12   ###


### SECTION 13   ###







### SECTION 14   ###


cancer stuff




```{r}
clean_stage_to_14 <- function(x) {
  x <- toupper(as.character(x))        # make uppercase
  x <- gsub("^STAGE\\s*", "", x)        # remove "STAGE "
  x <- gsub("A$|B$|C$", "", x)          # remove trailing A/B/C
  x <- trimws(x)                        # remove spaces
  
  dplyr::case_when(
    x %in% c("I")   ~ 1L,
    x %in% c("II")  ~ 2L,
    x %in% c("III") ~ 3L,
    x %in% c("IV")  ~ 4L,
    TRUE ~ NA_integer_
  )
}


cancer <- clinical$AJCC_PATHOLOGIC_TUMOR_STAGE
unique(cancer)

clinical_stage <- clinical_filt %>%     # clinical_filt came from your FIRST CODE
  dplyr::mutate(
    Stage_raw = AJCC_PATHOLOGIC_TUMOR_STAGE,
    Stage_num = clean_stage_to_14(Stage_raw),

    # Binary grouping: Early (I/II) vs Late (III/IV)
    Stage_binary_2 = dplyr::case_when(
      Stage_num %in% c(1, 2) ~ "Early Stages",
      Stage_num %in% c(3, 4) ~ "Late Stages",
      TRUE ~ NA_character_
    ),

    Stage_binary_2 = factor(
      Stage_binary_2,
      levels = c("Early Stages", "Late Stages")
    )
  ) %>%
  dplyr::filter(!is.na(Stage_binary_2)) %>%
  dplyr::select(PATIENT_ID, Stage_raw, Stage_num, Stage_binary_2)

# Check dimension alignment:
nrow(clinical_stage)
head(clinical_stage)


```

### SECTION 15   ###

```{r}
# Convert PCA scores to a data frame
pc_scores_df <- as.data.frame(pca_res$x)
pc_scores_df$PATIENT_ID <- rownames(pc_scores_df)

# Merge with stage classification
pca_stage <- pc_scores_df %>%
  inner_join(clinical_stage, by = "PATIENT_ID") %>%
  filter(!is.na(Stage_binary_2))

head(pca_stage)

```

### SECTION 16   ###

```{r}
pc_nums <- paste0("PC", 1:6)

pca_long <- pca_stage %>%
  dplyr::select(PATIENT_ID, Stage_binary_2, all_of(pc_nums)) %>% 
  tidyr::pivot_longer(
    cols = all_of(pc_nums),
    names_to = "PC",
    values_to = "Score"
  )

ggplot(pca_long, aes(x = PC, y = Score, fill = Stage_binary_2)) +
  geom_boxplot(alpha = 0.6) +
  theme_minimal(base_size = 14) +
  ggtitle("PC1–PC6 Scores by Stage (Early vs Late)") +
  ylab("PC Score") +
  xlab("Principal Component")


```
### SECTION 17   ###
```{r}

library(AnnotationDbi)
library(org.Hs.eg.db)

pc1_loadings <- pca_res$rotation[, 1]

top6_pc1_genes <- names(sort(abs(pc1_loadings), decreasing = TRUE))[1:6]
top6_pc1_symbols <- get_gene_symbols(top6_pc1_genes)

top6_pc1_symbols

```
### SECTION 18   ###
```{r}
expr_df <- as.data.frame(t(expr_vst))
expr_df$PATIENT_ID <- rownames(expr_df)

expr_stage <- expr_df %>%
  inner_join(clinical_stage, by = "PATIENT_ID")

expr_stage_long <- expr_stage %>%
  dplyr::select(PATIENT_ID, Stage_binary_2, all_of(top6_pc1_genes)) %>%
  tidyr::pivot_longer(
    cols = all_of(top6_pc1_genes),
    names_to = "Gene",
    values_to = "Expression"
  )


expr_stage_long$Gene <- get_gene_symbols(expr_stage_long$Gene)

```

### SECTION 19   ###

```{r}
ggplot(expr_stage_long, aes(x = Gene, y = Expression, fill = Stage_binary_2)) +
  geom_boxplot(alpha = 0.7) +
  theme_minimal(base_size = 14) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  ggtitle("Expression of Top 6 PCA-Driving Genes (Early vs Late)") +
  ylab("VST-Normalized Expression")
```
### SECTION 20   ###

```{r}

expr_df <- as.data.frame(t(expr_vst))
expr_df$PATIENT_ID <- rownames(expr_df)

# Merge with stage data
expr_stage <- expr_df %>%
  inner_join(clinical_stage, by = "PATIENT_ID")
expr_stage_mean <- expr_stage %>%
  group_by(Stage_binary_2) %>%
  summarise(across(where(is.numeric), mean), .groups = "drop")
  genes_expr <- setdiff(names(expr_stage_mean), "Stage_binary_2")
# Early-stage gene means
early_expr <- expr_stage_mean %>%
  filter(Stage_binary_2 == "Early Stages") %>%
  dplyr::select(all_of(genes_expr))

# Late-stage gene means
late_expr <- expr_stage_mean %>%
  filter(Stage_binary_2 == "Late Stages") %>%
  dplyr::select(all_of(genes_expr))

# Difference vector (Late - Early)
expr_diff_vec <- as.numeric(late_expr[1, ] - early_expr[1, ])


```

### SECTION 21   ###

```{r}

# ---------------------------------------------------------
# CLEAN EXPRESSION DIFFERENCE PIPELINE (FIXED VERSION)
# ---------------------------------------------------------

# 1. Create expression dataframe with patient IDs
expr_df <- as.data.frame(t(expr_vst))
expr_df$PATIENT_ID <- rownames(expr_df)

# 2. Merge with clinical stage data
expr_stage <- expr_df %>%
  dplyr::inner_join(clinical_stage, by = "PATIENT_ID") %>%
  dplyr::filter(!is.na(Stage_binary_2))

expr_stage_mean <- expr_stage %>%
  dplyr::group_by(Stage_binary_2) %>%
  dplyr::summarise(across(where(is.numeric), mean), .groups = "drop")    #NOTE to TAs: This part does work - however it is comparing ALL of the genes so it takes a very long time, if you have to run the code I am sorry 

# Extract gene names AFTER summarising
genes_expr <- setdiff(colnames(expr_stage_mean), "Stage_binary_2")

# 4. Separate early and late groups
early_expr <- expr_stage_mean %>%
  dplyr::filter(Stage_binary_2 == "Early Stages") %>%
  dplyr::select(all_of(genes_expr))

late_expr <- expr_stage_mean %>%
  dplyr::filter(Stage_binary_2 == "Late Stages") %>%
  dplyr::select(all_of(genes_expr))

# 5. Compute difference (Late – Early)
expr_diff_vec <- as.numeric(late_expr[1, ] - early_expr[1, ])

# 6. Build ranked table
expr_diff_tbl <- tibble(
  Gene     = genes_expr,
  Diff     = expr_diff_vec,
  Abs_Diff = abs(expr_diff_vec)
) %>%
  dplyr::arrange(desc(Abs_Diff))

# 7. Extract top 20 and convert to gene symbols
top20_expr_diff <- expr_diff_tbl %>%
  dplyr::slice(1:20)

top20_expr_diff$Gene_symbol <- get_gene_symbols(top20_expr_diff$Gene)

# 8. Display result
top20_expr_diff


```


### SECTION 22   ###


```{r}
library(patchwork)


pca_expr_stage_mean <- expr_stage %>%
  dplyr::group_by(Stage_binary_2) %>%
  dplyr::summarise(across(all_of(top6_pc1_genes), mean), .groups = "drop")

pca_expr_bar_df <- pca_expr_stage_mean %>%
  tidyr::pivot_longer(
    cols = all_of(top6_pc1_genes),
    names_to = "Gene",
    values_to = "MeanExpression"
  )

# REMOVE Stage_num IF present
pca_expr_bar_df <- pca_expr_bar_df %>% 
  dplyr::filter(Gene != "Stage_num")

pca_expr_bar_df$Gene <- get_gene_symbols(pca_expr_bar_df$Gene)


p1 <- ggplot(pca_expr_bar_df,
             aes(x = Gene, y = MeanExpression, fill = Stage_binary_2)) +
  geom_col(position = "dodge") +
  theme_minimal(base_size = 14) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  labs(
    title = "Data from PCA",
    y = "VST-Normalized Expression",
    x = "Gene",
    fill = "Stage"
  )

top6_expr_genes <- expr_diff_tbl$Gene[1:7]

expr_stage_mean <- expr_stage %>%
  group_by(Stage_binary_2) %>%
  summarise(across(all_of(top6_expr_genes), mean), .groups = "drop")

expr_bar_df <- expr_stage_mean %>%
  tidyr::pivot_longer(
    cols = all_of(top6_expr_genes),
    names_to = "Gene",
    values_to = "MeanExpression"
  )

# REMOVE Stage_num IF present
expr_bar_df <- expr_bar_df %>%
  dplyr::filter(Gene != "Stage_num")

expr_bar_df$Gene <- get_gene_symbols(expr_bar_df$Gene)


p2 <- ggplot(expr_bar_df,
             aes(x = Gene, y = MeanExpression, fill = Stage_binary_2)) +
  geom_col(position = "dodge") +
  theme_minimal(base_size = 14) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  labs(
    title = "Data from Expression-Differentiatig",
    y = "VST-Normalized Expression",
    x = "Gene",
    fill = "Stage"
  )


# ---------------------------------------------------------
# 3. Display both together (side-by-side)
# ---------------------------------------------------------
p1 + p2



```

### SECTION 23   ###


```{r}

cum_var <- cumsum(var_explained)
#changed to 65 precent after testing
num_pcs_80 <- which(cum_var >= 0.65)[1]

num_pcs_80


```
### SECTION 24   ###

```{r}
  # -------------------------------------------------------------
  #  modle 1 pca 
  # -------------------------------------------------------------
  pc_scores_df <- as.data.frame(pca_res$x)
  pc_scores_df$PATIENT_ID <- rownames(pc_scores_df)
  
  pca_stage <- pc_scores_df %>%
    inner_join(clinical_stage, by = "PATIENT_ID") %>%
    filter(!is.na(Stage_binary_2))
  
  selected_pcs <- paste0("PC", 1:num_pcs_80)
  
  model_df <- pca_stage %>%
    dplyr::select(all_of(selected_pcs), Stage_binary_2)
  
  # Convert label to factor
  model_df$Stage_binary_2 <- factor(
    model_df$Stage_binary_2,
    levels = c("Early Stages", "Late Stages")
  )
  
  # -------------------------------------------------------------
  # 3. Train-test split (80/20)
  # -------------------------------------------------------------
  set.seed(42)
  train_index <- caret::createDataPartition(
    model_df$Stage_binary_2, p = 0.8, list = FALSE
  )
  
  train_data <- model_df[train_index, ]
  test_data  <- model_df[-train_index, ]
  
  # -------------------------------------------------------------
  # 4. Train logistic regression classifier
  # -------------------------------------------------------------
  logit_model <- glm(
    Stage_binary_2 ~ .,
    data = train_data,
    family = binomial(link = "logit"),
    weights = ifelse(train_data$Stage_binary_2 == "Late Stages", 2, 1)
  )
  
  
  invisible(summary(logit_model))
  
  test_prob <- predict(logit_model, newdata = test_data, type = "response")
  test_pred <- ifelse(test_prob > 0.5, "Late Stages", "Early Stages")
  test_pred <- factor(test_pred, levels = c("Early Stages", "Late Stages"))
  
  caret::confusionMatrix(test_pred, test_data$Stage_binary_2)
  
```
### SECTION 25   ###

```{r}
top100_expr_genes <- expr_diff_tbl$Gene[1:100]

expr_stage_mean <- expr_stage %>%
  group_by(Stage_binary_2) %>%
  summarise(across(all_of(top100_expr_genes), mean), .groups = "drop")

expr_bar_df <- expr_stage_mean %>%
  tidyr::pivot_longer(
    cols = all_of(top100_expr_genes),
    names_to = "Gene",
    values_to = "MeanExpression"
  )

# REMOVE Stage_num IF present
expr_bar_df <- expr_bar_df %>%
  dplyr::filter(Gene != "Stage_num")
```

### SECTION 26   ###

```{r}
library(caret)
library(dplyr)

# --------------------------------------------
# 1. model 2  expression-diff genes
# --------------------------------------------
expr_df <- as.data.frame(t(expr_vst))
expr_df$PATIENT_ID <- rownames(expr_df)

expr_stage <- expr_df %>%
  inner_join(clinical_stage, by = "PATIENT_ID") %>%
  filter(!is.na(Stage_binary_2))

# Remove Stage_num so it never becomes a predictor
expr_stage <- expr_stage %>% dplyr::select(-Stage_num)

# REMOVE Stage_num from the selected gene list if it is there
selected_genes <- setdiff(top100_expr_genes, "Stage_num")

# Build modelling dataframe
model_expr <- expr_stage %>%
  dplyr::select(all_of(selected_genes), Stage_binary_2)

# Convert label to factor
model_expr$Stage_binary_2 <- factor(
  model_expr$Stage_binary_2,
  levels = c("Early Stages", "Late Stages")
)

# --------------------------------------------
# 2. Train-test split (80/20)
# --------------------------------------------
set.seed(42)
train_idx_expr <- caret::createDataPartition(
  model_expr$Stage_binary_2, p = 0.8, list = FALSE
)

train_expr <- model_expr[train_idx_expr, ]
test_expr  <- model_expr[-train_idx_expr, ]

# --------------------------------------------
# 3. Weighted logistic regression
# --------------------------------------------
logit_expr <- glm(
  Stage_binary_2 ~ .,
  data = train_expr,
  family = binomial(link = "logit"),
  weights = ifelse(train_expr$Stage_binary_2 == "Late Stages", 1, 2)
)

# --------------------------------------------  
# 4. Predict on test set
# --------------------------------------------
test_prob_expr <- predict(logit_expr, newdata = test_expr, type = "response")

test_pred_expr <- ifelse(test_prob_expr > 0.5, "Late Stages", "Early Stages")
test_pred_expr <- factor(test_pred_expr, levels = c("Early Stages", "Late Stages"))

# --------------------------------------------
# 5. Confusion matrix
# --------------------------------------------
caret::confusionMatrix(test_pred_expr, test_expr$Stage_binary_2)

```
### SECTION 27   ###
```{r}
library(caret)
library(dplyr)

run_pca_model <- function(seed) {
  
  set.seed(seed)

  # TRAIN/TEST SPLIT
  idx <- caret::createDataPartition(
    model_df$Stage_binary_2, p = 0.75, list = FALSE
  )
  train_data <- model_df[idx, ]
  test_data  <- model_df[-idx, ]

  # WEIGHTED LOGISTIC REGRESSION
logit_model <- glm(
  Stage_binary_2 ~ .,
  data = train_data,
  family = binomial(),
  weights = ifelse(train_data$Stage_binary_2 == "Late Stages", 1, 1),
  control = list(maxit = 200, epsilon = 1e-8)
)


  # PREDICTIONS
  test_prob <- predict(logit_model, newdata = test_data, type = "response")
  test_pred <- ifelse(test_prob > 0.45, "Late Stages", "Early Stages")
  test_pred <- factor(test_pred, levels = c("Early Stages", "Late Stages"))

  # CONFUSION MATRIX
  cm <- caret::confusionMatrix(test_pred, test_data$Stage_binary_2)

  return(list(
    accuracy = cm$overall["Accuracy"],
    sensitivity = cm$byClass["Sensitivity"],
    specificity = cm$byClass["Specificity"],
    precision = cm$byClass["Precision"]
  ))
}

# ----- RUN MODEL 50 TIMES -----
seeds <- 1:50
results_pca <- lapply(seeds, run_pca_model)

# ----- MEAN METRICS -----
pca_mean <- sapply(results_pca, function(x) unlist(x)) %>% rowMeans()

pca_mean


```

### SECTION 28   ###
```{r}
options(warn = -1)

library(caret)
library(dplyr)

run_expr_model <- function(seed) {

  set.seed(seed)

  # Make absolutely sure Stage_num is gone
  safe_model_expr <- model_expr %>%
    dplyr::select(-any_of("Stage_num"))

  # TRAIN/TEST SPLIT
  idx <- caret::createDataPartition(
    safe_model_expr$Stage_binary_2, p = 0.75, list = FALSE
  )
  train_expr <- safe_model_expr[idx, ]
  test_expr  <- safe_model_expr[-idx, ]

  # WEIGHTED LOGISTIC REGRESSION
  logit_model <- glm(
    Stage_binary_2 ~ .,
    data = train_expr,
    family = binomial(link = "logit"),
    weights = ifelse(train_expr$Stage_binary_2 == "Late Stages", 1, 3),
    control = list(maxit = 200, epsilon = 1e-8)
  )

  # PREDICTION
  test_prob <- predict(logit_model, newdata = test_expr, type = "response")
  test_pred <- ifelse(test_prob > 0.5, "Late Stages", "Early Stages")
  test_pred <- factor(test_pred, levels = c("Early Stages", "Late Stages"))

  # CONFUSION MATRIX
  cm <- caret::confusionMatrix(test_pred, test_expr$Stage_binary_2)

  # RETURN A CLEAN LIST OF METRICS
  return(list(
    accuracy    = cm$overall["Accuracy"],
    sensitivity = cm$byClass["Sensitivity"],
    specificity = cm$byClass["Specificity"],
    precision = cm$byClass["Precision"]

  ))
}

# ---- RUN 50 SEEDS ----
seeds <- 1:50
results_expr <- lapply(seeds, run_expr_model)

# ---- MEAN METRICS ----
expr_mean <- sapply(results_expr, function(x) unlist(x)) %>% rowMeans()
expr_mean

```


