---
title: "Sarah"
author: "Sarah Dumont"
date: "2025-11-25"
output: html_document
---


### SECTION: 1 ###


```{r}
#loaded Libraries 

library(dplyr)
library(readr)
library(tibble)
library(ggplot2)
library(tidyr)
library(e1071)
library(purrr)

```



### SECTION: 2 ###



```{r setup, message=FALSE, warning=FALSE}
knitr::opts_chunk$set(echo = TRUE)

clinical <- read_tsv(
  "data_clinical_patient.txt",
  comment = "#",
  show_col_types = FALSE
)

mutation <- read.delim(
  "data_mutations.txt",
  sep = "\t",
  header = TRUE,
  stringsAsFactors = FALSE
)

# --- Helper: normalize TCGA IDs to first 12 chars with dashes ---
normalize_tcga <- function(x) {
  x <- as.character(x)
  x <- gsub("\\.", "-", x)
  substr(x, 1, 12)
}

# --- Get patient IDs in clinical and mutation data ---
clin_ids <- unique(normalize_tcga(clinical$PATIENT_ID))
mut_ids  <- unique(normalize_tcga(mutation$Tumor_Sample_Barcode))

# Only keep patients that are in BOTH clinical and mutation datasets
common_ids <- intersect(clin_ids, mut_ids)

# --- Filter clinical to common patients (not actually used below, but kept for completeness) ---
clinical <- clinical %>%
  mutate(PATIENT_ID = normalize_tcga(PATIENT_ID)) %>%
  filter(PATIENT_ID %in% common_ids)

# --- Filter to NON-SYNONYMOUS mutations (single clean list) ---
nonsynonymous <- c(
  "Missense_Mutation", "Nonsense_Mutation",
  "Nonstop_Mutation",
  "Frame_Shift_Del", "Frame_Shift_Ins",
  "In_Frame_Del", "In_Frame_Ins",
  "Splice_Site", "Translation_Start_Site"
)

mutation_ns <- mutation %>%
  mutate(PATIENT_ID = normalize_tcga(Tumor_Sample_Barcode)) %>%
  filter(
    Variant_Classification %in% nonsynonymous,
    PATIENT_ID %in% common_ids
  )

# --- Build Gene × Patient mutation matrix (0/1) ---
genes    <- sort(unique(mutation_ns$Hugo_Symbol))
patients <- sort(unique(mutation_ns$PATIENT_ID))

mut_matrix <- matrix(
  0L,
  nrow = length(genes),
  ncol = length(patients),
  dimnames = list(genes, patients)
)

for (i in seq_len(nrow(mutation_ns))) {
  g <- mutation_ns$Hugo_Symbol[i]
  p <- mutation_ns$PATIENT_ID[i]
  mut_matrix[g, p] <- 1L
}

# --- Gene mutation frequencies across the cohort ---
gene_freq <- rowSums(mut_matrix)  # how many patients mutated per gene

top_n <- 20  # change this if you want more/less
top_genes <- sort(gene_freq, decreasing = TRUE)[1:top_n]

top_genes_df <- tibble(
  Gene = names(top_genes),
  Mutated_Patients = as.integer(top_genes)
)

```





### SECTION: 3 ###






```{r}
# Bar plot of most frequently mutated genes ---
ggplot(top_genes_df,
       aes(x = reorder(Gene, Mutated_Patients),
           y = Mutated_Patients)) +
  geom_col(fill = "steelblue") +
  coord_flip() +
  theme_minimal(base_size = 14) +
  labs(
    title = paste("Top", top_n, "Most Frequently Mutated Genes"),
    x = "Gene",
    y = "Number of Patients with a Non-synonymous Mutation"
  )
```




### SECTION: 4 ###




```{r}

gene_mut_count <- rowSums(mut_matrix)
gene_df <- data.frame(
  Gene = names(gene_mut_count),
  Mutated_Patients = gene_mut_count
)


tmb_per_patient <- colSums(mut_matrix)
tmb_df <- data.frame(
  Patient = names(tmb_per_patient),
  TMB = tmb_per_patient
)

# Plot: TMB distribution
ggplot(tmb_df, aes(x = TMB)) +
  geom_histogram(binwidth = 1, fill = "purple", color = "black") +
  geom_vline(aes(xintercept = mean(TMB)), color = "red", linetype = "dashed", size = 1) +
  labs(title = "Tumor Mutational Burden (TMB) Distribution",
       x = "Mutations per Patient (TMB)",
       y = "Number of Patients") +
  theme_minimal()

```


### SECTION: 5 ###
  
```{r}
patient_mut_counts <- colSums(mut_matrix)

mean_mutated_genes <- mean(patient_mut_counts)

mean_mutated_genes

```



### SECTION: 6 ###



```{r}
gene_variance <- apply(mut_matrix, 1, var)

# Sort descending
gene_variance_sorted <- sort(gene_variance, decreasing = TRUE)

# Look at the top variance genes
head(gene_variance_sorted, 20)

```

### SECTION: 7 ###


```{r}

# 1. Gene variance across patients
gene_variance <- apply(mut_matrix, 1, var)

# 2. Sort by variance (high → low)
gene_variance_sorted <- sort(gene_variance, decreasing = TRUE)

# 3. Basic variance curve (log scale)
plot(
  gene_variance_sorted,
  type = "l",
  log = "y",
  lwd = 2,
  col = "blue",
  xlab = "Gene Rank (sorted by variance)",
  ylab = "Variance (log scale)",
  main = "Gene Variance Distribution with Elbow"
)

# 4. Approximate elbow via second derivative on log-variance
y_log <- log10(gene_variance_sorted + 1e-10)  # avoid log(0)
d1 <- diff(y_log)
d2 <- diff(d1)

# We look for the point of **maximum curvature**
elbow <- which.max(-d2) + 1   # +1 to align with original index

elbow

# 5. Add vertical line at elbow
abline(v = elbow, col = "red", lwd = 2, lty = 2)

text(
  x = elbow,
  y = max(gene_variance_sorted),
  labels = paste("Elbow =", elbow),
  pos = 4,
  col = "red"
)

# 6. Extract "significant" genes (above elbow)
significant_genes <- names(gene_variance_sorted)[1:elbow]
head(significant_genes)
length(significant_genes)



```

✔ A small group of genes drive almost all variation ✔ Most genes in the genome are not mutated in this cancer type ✔ The “important” high-variance genes are the ones involved in the disease

Top 6 most frequelntly mutated genes "TP53" "TTN" "CTNNB1" "MUC16" "ALB" "PCLO"



### SECTION: 8 ###

```{r}
# --- PCA to Assess Variance Structure ---

# 1. Transpose mut_matrix → rows = patients, cols = genes
X <- t(mut_matrix)

# 2. PCA (center & scale recommended for binary mutation data)
pca_res <- prcomp(X, center = TRUE, scale. = TRUE)

# 3. Variance explained per PC
var_explained <- pca_res$sdev^2 / sum(pca_res$sdev^2)

# 4. Cumulative variance explained
cum_var <- cumsum(var_explained)

# 5. Find how many PCs needed for 90% variance
num_pcs_90 <- which(cum_var >= 0.90)[1]

num_pcs_90

```

I ran PCA here to understand why we don't want to be using PCA! 270 PCAs is alot typically when doing PCA analysis we end up / want to end up with less then 10 PCAs, however here the reason that we have so many is because there is so many genes and some of them only have a couple mutations means that this will have low variance but becasue so many of them have low varianvce PCA needs alot more PCAs to achive 90 percent


### SECTION: 9 ###

Question 2: can a model be predicted that can accuratly determine the stage of cancer.

```{r}
stage_col <- "AJCC_PATHOLOGIC_TUMOR_STAGE"

                            

clean_stage_to_14 <- function(x) {
  x <- toupper(as.character(x))
  x <- gsub("^STAGE\\s*", "", x)  
  x <- gsub("[A]+$", "", x)     
  x <- trimws(x)

  dplyr::case_when(
    x %in% c("I", "1")   ~ 1L,
    x %in% c("II", "2")  ~ 2L,
    x %in% c("III", "3") ~ 3L,
    x %in% c("IV", "4")  ~ 4L,
    TRUE ~ NA_integer_
  )
}

clinical_stage <- clinical %>%
  mutate(
    Stage_raw = .data[[stage_col]],
    Stage_num = clean_stage_to_14(Stage_raw),
    Stage_grp = factor(
      Stage_num,
      levels = 1:4,
      labels = paste("Stage", 1:4)
    )
  ) %>%
  filter(!is.na(Stage_num)) %>%
  select(PATIENT_ID, Stage_raw, Stage_num, Stage_grp)

clinical_stage

```



### SECTION: 10 ###




```{r}

tmb_df <- tmb_df %>%
  rename(PATIENT_ID = Patient)

tmb_stage <- tmb_df %>%
  inner_join(clinical_stage, by = "PATIENT_ID")

ggplot(tmb_stage, aes(x = Stage_grp, y = TMB)) +
  geom_boxplot(fill = "lightblue", alpha = 0.5) +
  geom_jitter(width = 0.2, alpha = 0.6) +
  theme_minimal(base_size = 14) +
  labs(
    title = "Tumor Mutation Burden by Collapsed Stage (I–IV)",
    x = "Stage (Grouped)",
    y = "TMB (Non-synonymous Mutations)"
  )

```

Look into the TMB to see if maybe total number of mutations could be an indicator of stage of cancer. From this box plot it looks like most are in similar ranges meaning purely quantity of mutated genes is not a good indicator. So further analysis on specifc mutated genes would potenitally be a better indicator of cancer stage be a better indicator

TMB alone is not a good staging biomarker These box plots do not have enough variation a model will not be able to distiguish these catigoeies



### SECTION: 11 ###


```{r}
table(clinical_stage$Stage_grp)

stage_counts <- clinical_stage %>%
  count(Stage_grp)

stage_counts

```

Troubles: cannot do with by simply splitting stages because stage for only has 4 patients so we will combine into early and late stage cancer instead.



### SECTION: 12 ###


```{r}

clinical_stage <- clinical_stage %>%
  mutate(
    Stage_binary_2 = case_when(
      Stage_num %in% c(1, 2) ~ "Early Stages",
      Stage_num %in% c(3, 4) ~ "Late Stages",
      TRUE ~ NA_character_
    ),
    Stage_binary_2 = factor(
      Stage_binary_2,
      levels = c("Early Stages", "Late Stages")
    )
  )

```



### SECTION: 13 ###

```{r}
gene_freq <- rowSums(mut_matrix)
top_genes <- names(sort(gene_freq, decreasing = TRUE))[1:100] # look at top 100 mutated genes
top_genes

gene_df <- t(mut_matrix[top_genes, ]) %>%
  as.data.frame() %>%
  mutate(PATIENT_ID = rownames(.)) %>%
  inner_join(clinical_stage, by = "PATIENT_ID")

# Calculate mutation proportion per stage
stage_gene_freq <- gene_df %>%
  group_by(Stage_binary_2) %>%
  summarise(across(top_genes, mean))

stage_gene_freq
```

### SECTION: 14 ###


```{r}

genes <- setdiff(names(stage_gene_freq), "Stage_binary_2")

# Extract rows for Early and Late
early_row <- stage_gene_freq %>%
  filter(Stage_binary_2 == "Early Stages") %>%
  select(all_of(genes))

late_row <- stage_gene_freq %>%
  filter(Stage_binary_2 == "Late Stages") %>%
  select(all_of(genes))

diff_vec <- as.numeric(late_row[1, ] - early_row[1, ])

gene_diff_tbl <- tibble(
  Gene     = genes,
  Diff     = diff_vec,           
  Abs_Diff = abs(diff_vec)
) %>%
  arrange(desc(Abs_Diff))        

gene_diff_tbl <- gene_diff_tbl[1:20, ]

write.table(gene_diff_tbl, row.names = FALSE, sep = ",")

```



### SECTION: 15 ###



```{r}



gene_freq <- rowSums(mut_matrix)
top_50_genes <- names(sort(gene_freq, decreasing = TRUE))[1:100]     #changed to 100 but didnt want to rename

top50_rank_lookup <- setNames(seq_along(top_50_genes), top_50_genes)

gene_diff_ranked <- gene_diff_tbl %>%
  mutate(
    DiffRank = row_number(),                 
    Rank_in_Top50 = top50_rank_lookup[Gene] 
  )

gene_diff_ranked

```

### SECTION: 16 ###



```{r}
top10_genes <- gene_diff_tbl$Gene[1:6]

bar_df <- stage_gene_freq %>%
  select(Stage_binary_2, all_of(top10_genes)) %>%
  pivot_longer(cols = all_of(top10_genes),
               names_to = "Gene",
               values_to = "Proportion")

ggplot(bar_df, aes(x = Gene, y = Proportion, fill = Stage_binary_2)) +
  geom_col(position = "dodge") +
  theme_minimal(base_size = 14) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  labs(
    title = "Mutation Proportions of Top 6 Most Mutated Genes",
    y = "Proportion Mutated",
    fill = "Stage" 
  )

```


### SECTION: 17 ###


```{r}
gene_freq <- rowSums(mut_matrix)
top10_genes <- names(sort(gene_freq, decreasing = TRUE))[1:6]

bar_df <- stage_gene_freq %>%
  select(Stage_binary_2, all_of(top10_genes)) %>%
  pivot_longer(cols = all_of(top10_genes),
               names_to = "Gene",
               values_to = "Proportion")

ggplot(bar_df, aes(x = Gene, y = Proportion, fill = Stage_binary_2)) +
  geom_col(position = "dodge") +
  theme_minimal(base_size = 14) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  labs(
    title = "Mutation Proportions of Top 6 Differentiating Genes",
    y = "Proportion Mutated",
    fill = "Stage" 
  )
```


This shows that after performing some preliminary re-organization of the data, we now have a structured set of mutation features that can more effectively separate cancer stages into early and late groups. This improves the usefulness of the dataset and should increase the accuracy of any future predictive models by ensuring they are trained on meaningful, relevant information — garbage in, garbage out.

Additionally, we can now see that DNAH6 appears only in the early-stage group. This suggests that DNAH6 may be a particularly valuable feature for helping the model distinguish between early and late cancer stages.


### SECTION: 18 ###

```{r}
clean_stage_to_14 <- function(x) {
  x <- toupper(as.character(x))
  x <- gsub("^STAGE\\s*", "", x)
  x <- gsub("[A]+$", "", x)
  x <- trimws(x)

  dplyr::case_when(
    x %in% c("I", "1")   ~ 1L,
    x %in% c("II", "2")  ~ 2L,
    x %in% c("III", "3") ~ 3L,
    x %in% c("IV", "4")  ~ 4L,
    TRUE ~ NA_integer_
  )
}

clinical_stage <- clinical %>%
  mutate(
    Stage_raw = AJCC_PATHOLOGIC_TUMOR_STAGE,
    Stage_num = clean_stage_to_14(Stage_raw),
    Stage_lowhigh = case_when(
      Stage_num %in% c(1, 2) ~ "Low Stage",   # Stage I–II
      Stage_num %in% c(3, 4) ~ "High Stage",  # Stage III–IV
      TRUE ~ NA_character_
    )
  ) %>%
  filter(!is.na(Stage_lowhigh)) %>%
  select(PATIENT_ID, Stage_num, Stage_lowhigh)
```





### SECTION: 19 ###





```{r}
library(dplyr)
library(e1071)

genes_for_model <- intersect(gene_diff_tbl$Gene[1:30], rownames(mut_matrix))

mut_df <- t(mut_matrix[genes_for_model, , drop = FALSE]) %>%
  as.data.frame() %>%
  mutate(PATIENT_ID = rownames(.))

df <- mut_df %>%
  inner_join(clinical_stage %>% select(PATIENT_ID, Stage_lowhigh),
             by = "PATIENT_ID") %>%
  filter(!is.na(Stage_lowhigh))

df$Stage_lowhigh <- factor(df$Stage_lowhigh)

# Prepare results storage
results <- data.frame(seed = integer(), accuracy = numeric(),
                      sensitivity = numeric(), specificity = numeric(),
                      precision = numeric(), stringsAsFactors = FALSE)

threshold <- 0.99

# Run multiple seeds
for(i in 1:10){   # can increase to 50
  set.seed(i)
  
  # stratified split
  low <- df %>% filter(Stage_lowhigh == "Low Stage")
  high <- df %>% filter(Stage_lowhigh == "High Stage")
  
  # skip if too few patients
  if(nrow(low) < 2 | nrow(high) < 2) next
  
  low_idx <- sample(seq_len(nrow(low)), size = floor(0.75 * nrow(low)))
  high_idx <- sample(seq_len(nrow(high)), size = floor(0.75 * nrow(high)))
  
  train <- bind_rows(low[low_idx, ], high[high_idx, ])
  test  <- bind_rows(low[-low_idx, ], high[-high_idx, ])
  
  train_nb <- train[, c(genes_for_model, "Stage_lowhigh")]
  
  nb_model <- naiveBayes(Stage_lowhigh ~ ., data = train_nb)
  
  # Predict probabilities
  nb_prob <- predict(nb_model, newdata = test[, genes_for_model, drop = FALSE], type = "raw")
  
  # Handle cases where one class is missing
  if(!"High Stage" %in% colnames(nb_prob)){
    nb_prob <- cbind(nb_prob, "High Stage" = 0)
  }
  
  nb_pred <- ifelse(nb_prob[, "High Stage"] > threshold, "High Stage", "Low Stage")
  nb_pred <- factor(nb_pred, levels = levels(df$Stage_lowhigh))
  
  cm <- table(Predicted = nb_pred, Actual = test$Stage_lowhigh)
  
  TP <- cm["High Stage", "High Stage"] %||% 0
  TN <- cm["Low Stage",  "Low Stage"]  %||% 0
  FP <- cm["High Stage", "Low Stage"]  %||% 0
  FN <- cm["Low Stage",  "High Stage"] %||% 0
  
  results <- rbind(results, data.frame(
    seed = i,
    accuracy = (TP + TN)/(TP+TN+FP+FN),
    sensitivity = if((TP+FN)>0) TP/(TP+FN) else NA,
    specificity = if((TN+FP)>0) TN/(TN+FP) else NA,
    precision = if((TP+FP)>0) TP/(TP+FP) else NA
  ))
}

# Average metrics
colMeans(results[,-1], na.rm = TRUE)

```
Because Naive Bayes tends to produce extremely high or low probabilities on this sparse mutation data, a very high decision threshold (0.999) was used for calling a tumor “High Stage”. This biases the classifier toward catching almost all truly high-stage tumors (sensitivity ≈ 94%), but at the cost of low specificity (≈ 32%) and modest overall accuracy (≈ 46%).


Could be a valid model because it rarely misses the high 

### SECTION: 20 ###



```{r}
library(dplyr)
library(e1071)

# --------------------------------
# 1. Gene list for the model
# --------------------------------

gene_freq <- rowSums(mut_matrix)
genes_for_model_2 <- names(sort(gene_freq, decreasing = TRUE))[1:30]

# ensure genes exist in mut_matrix
genes_for_model_2 <- intersect(genes_for_model_2, rownames(mut_matrix))

# --------------------------------
# 2. Build patient × gene + stage data
# --------------------------------

mut_df <- t(mut_matrix[genes_for_model_2, , drop = FALSE]) %>%
  as.data.frame() %>%
  mutate(PATIENT_ID = rownames(.))

df <- mut_df %>%
  inner_join(
    clinical_stage %>% select(PATIENT_ID, Stage_lowhigh),
    by = "PATIENT_ID"
  ) %>%
  filter(!is.na(Stage_lowhigh))

df$Stage_lowhigh <- factor(df$Stage_lowhigh)

# --------------------------------
# 3. Run multiple seeds with stratified splits
# --------------------------------

threshold <- 0.75
results <- data.frame(seed = integer(),
                      accuracy = numeric(),
                      sensitivity = numeric(),
                      specificity = numeric(),
                      precision = numeric(),
                      stringsAsFactors = FALSE)

`%||%` <- function(x, y) if (is.null(x) || length(x) == 0) y else x

for(i in 1:10){   # change to 50 if desired
  set.seed(i)
  
  low <- df %>% filter(Stage_lowhigh == "Low Stage")
  high <- df %>% filter(Stage_lowhigh == "High Stage")
  
  # skip iteration if too few patients in any class
  if(nrow(low) < 2 | nrow(high) < 2) next
  
  low_idx <- sample(seq_len(nrow(low)), size = floor(0.75 * nrow(low)))
  high_idx <- sample(seq_len(nrow(high)), size = floor(0.75 * nrow(high)))
  
  train <- bind_rows(low[low_idx, ], high[high_idx, ])
  test  <- bind_rows(low[-low_idx, ], high[-high_idx, ])
  
  train_nb <- train[, c(genes_for_model_2, "Stage_lowhigh")]
  
  nb_model <- naiveBayes(Stage_lowhigh ~ ., data = train_nb)
  
  nb_prob <- predict(nb_model, newdata = test[, genes_for_model_2, drop = FALSE], type = "raw")
  
  # handle missing High Stage column
  if(!"High Stage" %in% colnames(nb_prob)){
    nb_prob <- cbind(nb_prob, "High Stage" = 0)
  }
  
  nb_pred <- ifelse(nb_prob[, "High Stage"] > threshold, "High Stage", "Low Stage")
  nb_pred <- factor(nb_pred, levels = levels(df$Stage_lowhigh))
  
  cm <- table(Predicted = nb_pred, Actual = test$Stage_lowhigh)
  
  TP <- cm["High Stage", "High Stage"] %||% 0
  TN <- cm["Low Stage",  "Low Stage"]  %||% 0
  FP <- cm["High Stage", "Low Stage"]  %||% 0
  FN <- cm["Low Stage",  "High Stage"] %||% 0
  
  results <- rbind(results, data.frame(
    seed = i,
    accuracy = (TP + TN)/(TP+TN+FP+FN),
    sensitivity = if((TP+FN)>0) TP/(TP+FN) else NA,
    specificity = if((TN+FP)>0) TN/(TN+FP) else NA,
    precision = if((TP+FP)>0) TP/(TP+FP) else NA
  ))
}

# --------------------------------
# 4. Average metrics across runs
# --------------------------------

summary_stats <- colMeans(results[,-1], na.rm = TRUE)
results
summary_stats

```

