---
title: "Sarah"
author: "Sarah Dumont"
date: "2025-11-25"
output: html_document
---

```{r setup, message=FALSE, warning=FALSE}
knitr::opts_chunk$set(echo = TRUE)

library(dplyr)
library(readr)
library(tibble)
library(ggplot2)

# --- Load Data ---
clinical <- read_tsv(
  "data_clinical_patient.txt",
  comment = "#",
  show_col_types = FALSE
)

mutation <- read.delim(
  "data_mutations.txt",
  sep = "\t",
  header = TRUE,
  stringsAsFactors = FALSE
)

# --- Helper: normalize TCGA IDs to first 12 chars with dashes ---
normalize_tcga <- function(x) {
  x <- as.character(x)
  x <- gsub("\\.", "-", x)
  substr(x, 1, 12)
}

# --- Get patient IDs in clinical and mutation data ---
clin_ids <- unique(normalize_tcga(clinical$PATIENT_ID))
mut_ids  <- unique(normalize_tcga(mutation$Tumor_Sample_Barcode))

# Only keep patients that are in BOTH clinical and mutation datasets
common_ids <- intersect(clin_ids, mut_ids)

# --- Filter clinical to common patients (not actually used below, but kept for completeness) ---
clinical <- clinical %>%
  mutate(PATIENT_ID = normalize_tcga(PATIENT_ID)) %>%
  filter(PATIENT_ID %in% common_ids)

# --- Filter to NON-SYNONYMOUS mutations (single clean list) ---
nonsynonymous <- c(
  "Missense_Mutation", "Nonsense_Mutation",
  "Nonstop_Mutation",
  "Frame_Shift_Del", "Frame_Shift_Ins",
  "In_Frame_Del", "In_Frame_Ins",
  "Splice_Site", "Translation_Start_Site"
)

mutation_ns <- mutation %>%
  mutate(PATIENT_ID = normalize_tcga(Tumor_Sample_Barcode)) %>%
  filter(
    Variant_Classification %in% nonsynonymous,
    PATIENT_ID %in% common_ids
  )

# --- Build Gene × Patient mutation matrix (0/1) ---
genes    <- sort(unique(mutation_ns$Hugo_Symbol))
patients <- sort(unique(mutation_ns$PATIENT_ID))

mut_matrix <- matrix(
  0L,
  nrow = length(genes),
  ncol = length(patients),
  dimnames = list(genes, patients)
)

for (i in seq_len(nrow(mutation_ns))) {
  g <- mutation_ns$Hugo_Symbol[i]
  p <- mutation_ns$PATIENT_ID[i]
  mut_matrix[g, p] <- 1L
}

# --- Gene mutation frequencies across the cohort ---
gene_freq <- rowSums(mut_matrix)  # how many patients mutated per gene

top_n <- 20  # change this if you want more/less
top_genes <- sort(gene_freq, decreasing = TRUE)[1:top_n]

top_genes_df <- tibble(
  Gene = names(top_genes),
  Mutated_Patients = as.integer(top_genes)
)



```

```{r}
# --- Bar plot of most frequently mutated genes ---
ggplot(top_genes_df,
       aes(x = reorder(Gene, Mutated_Patients),
           y = Mutated_Patients)) +
  geom_col(fill = "steelblue") +
  coord_flip() +
  theme_minimal(base_size = 14) +
  labs(
    title = paste("Top", top_n, "Most Frequently Mutated Genes"),
    x = "Gene",
    y = "Number of Patients with a Non-synonymous Mutation"
  )
```

```{r}
# --- Histogram of Mutation Counts per Gene ---

# mutation count for each gene
gene_mut_counts <- rowSums(mut_matrix)

# put into a data frame
gene_freq_df <- data.frame(
  Gene = names(gene_mut_counts),
  Count = as.numeric(gene_mut_counts)
)

# histogram
ggplot(gene_freq_df, aes(x = Count)) +
  geom_histogram(
    bins = 50,
    fill = "steelblue",
    color = "black",
    alpha = 0.7
  ) +
  theme_minimal(base_size = 14) +
  labs(
    title = "Distribution of Mutation Frequencies Across Genes",
    x = "Number of Patients Mutated (per gene)",
    y = "Number of Genes"
  )

```

```{r}
# Number of mutated genes per patient
patient_mut_counts <- colSums(mut_matrix)

# Mean TMB (mean mutated genes)
mean_mutated_genes <- mean(patient_mut_counts)

mean_mutated_genes

```

```{r}
# Gene variance across patients (binary 0/1)
gene_variance <- apply(mut_matrix, 1, var)

# Sort descending
gene_variance_sorted <- sort(gene_variance, decreasing = TRUE)

# Look at the top variance genes
head(gene_variance_sorted, 20)

```

```{r}

# 1. Gene variance across patients
gene_variance <- apply(mut_matrix, 1, var)

# 2. Sort by variance (high → low)
gene_variance_sorted <- sort(gene_variance, decreasing = TRUE)

# 3. Basic variance curve (log scale)
plot(
  gene_variance_sorted,
  type = "l",
  log = "y",
  lwd = 2,
  col = "blue",
  xlab = "Gene Rank (sorted by variance)",
  ylab = "Variance (log scale)",
  main = "Gene Variance Distribution with Elbow"
)

# 4. Approximate elbow via second derivative on log-variance
y_log <- log10(gene_variance_sorted + 1e-10)  # avoid log(0)
d1 <- diff(y_log)
d2 <- diff(d1)

# We look for the point of **maximum curvature**
elbow <- which.max(-d2) + 1   # +1 to align with original index

elbow

# 5. Add vertical line at elbow
abline(v = elbow, col = "red", lwd = 2, lty = 2)

text(
  x = elbow,
  y = max(gene_variance_sorted),
  labels = paste("Elbow =", elbow),
  pos = 4,
  col = "red"
)

# 6. Extract "significant" genes (above elbow)
significant_genes <- names(gene_variance_sorted)[1:elbow]
head(significant_genes)
length(significant_genes)



```

✔ A small group of genes drive almost all variation ✔ Most genes in the genome are not mutated in this cancer type ✔ The “important” high-variance genes are the ones involved in the disease

Top 6 most frequelntly mutated genes "TP53" "TTN" "CTNNB1" "MUC16" "ALB" "PCLO"

```{r}
# --- PCA to Assess Variance Structure ---

# 1. Transpose mut_matrix → rows = patients, cols = genes
X <- t(mut_matrix)

# 2. PCA (center & scale recommended for binary mutation data)
pca_res <- prcomp(X, center = TRUE, scale. = TRUE)

# 3. Variance explained per PC
var_explained <- pca_res$sdev^2 / sum(pca_res$sdev^2)

# 4. Cumulative variance explained
cum_var <- cumsum(var_explained)

# 5. Find how many PCs needed for 90% variance
num_pcs_90 <- which(cum_var >= 0.99)[1]

num_pcs_90

```

I ran PCA here to understand why we don't want to be using PCA! 270 PCAs is alot typically when doing PCA analysis we end up / want to end up with less then 10 PCAs, however here the reason that we have so many is because there is so many genes and some of them only have a couple mutations means that this will have low variance but becasue so many of them have low varianvce PCA needs alot more PCAs to achive 90 percent

```{r}
length(genes)
```

```{r}
gene_variance <- apply(mut_matrix, 1, var)
sort(gene_variance, decreasing = TRUE)[1:20]

```

Look into

Question 1:

Question 2: can a model be predicted that can accuratly determine the stage of cancer.

```{r}
# --- Find and collapse cancer stage into 1–4 ---
stage_col <- "AJCC_PATHOLOGIC_TUMOR_STAGE"

stage_col <- stage_candidates[1]     # use first match
stage_col                             # prints the chosen column

# 2) Function to clean stage strings
clean_stage_to_14 <- function(x) {
  x <- toupper(as.character(x))
  x <- gsub("^STAGE\\s*", "", x)   # remove "Stage "
  x <- gsub("[A]+$", "", x)      # remove letters A/B/C/etc.
  x <- trimws(x)

  dplyr::case_when(
    x %in% c("I", "1")   ~ 1L,
    x %in% c("II", "2")  ~ 2L,
    x %in% c("III", "3") ~ 3L,
    x %in% c("IV", "4")  ~ 4L,
    TRUE ~ NA_integer_
  )
}

# 3) Apply stage collapse
clinical_stage <- clinical %>%
  mutate(
    Stage_raw = .data[[stage_col]],
    Stage_num = clean_stage_to_14(Stage_raw),
    Stage_grp = factor(
      Stage_num,
      levels = 1:4,
      labels = paste("Stage", 1:4)
    )
  ) %>%
  filter(!is.na(Stage_num)) %>%
  select(PATIENT_ID, Stage_raw, Stage_num, Stage_grp)

clinical_stage


```

```{r}


tmb_stage <- tmb_df %>%
  inner_join(clinical_stage, by = "PATIENT_ID")

ggplot(tmb_stage, aes(x = Stage_grp, y = TMB)) +
  geom_boxplot(fill = "lightblue", alpha = 0.5) +
  geom_jitter(width = 0.2, alpha = 0.6) +
  theme_minimal(base_size = 14) +
  labs(
    title = "Tumor Mutation Burden by Collapsed Stage (I–IV)",
    x = "Stage (Grouped)",
    y = "TMB (Non-synonymous Mutations)"
  )

```

Look into the TMB to see if maybe total number of mutations could be an indicator of stage of cancer. From this box plot it looks like most are in similar ranges meaning purely quantity of mutated genes is not a good indicator. So further analysis on specifc mutated genes would potenitally be a better indicator of cancer stage be a better indicator

TMB alone is not a good staging biomarker These box plots do not have enough variation a model will not be able to distiguish these catigoeies

```{r}
# Count patients per collapsed stage
table(clinical_stage$Stage_grp)

# Or a nicer data frame
stage_counts <- clinical_stage %>%
  count(Stage_grp)

stage_counts

```

Troubles: cannot do with by simply splitting stages because stage for only has 4 patients so we will combine into early and late stage cancer instead.

```{r}

# Cobine into Early Stage and Late Stage
clinical_stage <- clinical_stage %>%
  mutate(
    Stage_binary_2 = case_when(
      Stage_num %in% c(1, 2) ~ "Early Stages",
      Stage_num %in% c(3, 4) ~ "Late Stages",
      TRUE ~ NA_character_
    ),
    Stage_binary_2 = factor(
      Stage_binary_2,
      levels = c("Early Stages", "Late Stages")
    )
  )

```

```{r}
gene_freq <- rowSums(mut_matrix)
top_genes <- names(sort(gene_freq, decreasing = TRUE))[1:100] # look at top 50 mutated genes
top_genes

# Build patient × gene table
gene_df <- t(mut_matrix[top_genes, ]) %>%
  as.data.frame() %>%
  mutate(PATIENT_ID = rownames(.)) %>%
  inner_join(clinical_stage, by = "PATIENT_ID")

# Calculate mutation proportion per stage
stage_gene_freq <- gene_df %>%
  group_by(Stage_binary_2) %>%
  summarise(across(top_genes, mean))

stage_gene_freq
#write.table(stage_gene_freq, row.names = FALSE, sep = ",")
```

```{r}

# Get just the gene columns (everything except Stage_binary_2)
genes <- setdiff(names(stage_gene_freq), "Stage_binary_2")

# Extract rows for Early and Late
early_row <- stage_gene_freq %>%
  filter(Stage_binary_2 == "Early Stages") %>%
  select(all_of(genes))

late_row <- stage_gene_freq %>%
  filter(Stage_binary_2 == "Late Stages") %>%
  select(all_of(genes))

# Compute Late - Early for each gene
diff_vec <- as.numeric(late_row[1, ] - early_row[1, ])

# Build a table: gene, difference, absolute difference
gene_diff_tbl <- tibble(
  Gene     = genes,
  Diff     = diff_vec,           # Late - Early (signed)
  Abs_Diff = abs(diff_vec)       # absolute difference
) %>%
  arrange(desc(Abs_Diff))         # sort by largest absolute difference

gene_diff_tbl <- gene_diff_tbl[1:10, ]

# If you want an easy copy/paste version:
write.table(gene_diff_tbl, row.names = FALSE, sep = ",")

```

```{r}



# --- Top 50 most mutated genes (from your mutation matrix) ---
gene_freq <- rowSums(mut_matrix)
top_50_genes <- names(sort(gene_freq, decreasing = TRUE))[1:100]

# lookup: gene -> rank 1..50 within top-50 list
top50_rank_lookup <- setNames(seq_along(top_50_genes), top_50_genes)

# --- Add ranks (1–10 by difference, + rank within top-50) ---
gene_diff_ranked <- gene_diff_tbl %>%
  mutate(
    DiffRank = row_number(),                 # 1..10 (already sorted)
    Rank_in_Top50 = top50_rank_lookup[Gene]  # 1..50 or NA if not in top 50
  )

gene_diff_ranked

```

```{r}
# Plot mutation proportion for Top 10 different genes
top10_genes <- gene_diff_tbl$Gene[1:10]

bar_df <- stage_gene_freq %>%
  select(Stage_binary_2, all_of(top10_genes)) %>%
  pivot_longer(cols = all_of(top10_genes),
               names_to = "Gene",
               values_to = "Proportion")

ggplot(bar_df, aes(x = Gene, y = Proportion, fill = Stage_binary_2)) +
  geom_col(position = "dodge") +
  theme_minimal(base_size = 14) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  labs(
    title = "Mutation Proportions of Top 4 Differentiating Genes",
    y = "Proportion Mutated"
  )

```

```{r}
# Step 1 — Take the top 25 highest-difference genes
top25_diff <- gene_diff_ranked %>%
  slice(1:30)

# Step 2 — Keep only genes that are ALSO within the top 20 most-mutated
top25_filtered <- top25_diff %>%
  filter(Rank_in_Top50 <= 30)

top25_filtered$Gene



```

This shows that after performing some preliminary re-organization of the data, we now have a structured set of mutation features that can more effectively separate cancer stages into early and late groups. This improves the usefulness of the dataset and should increase the accuracy of any future predictive models by ensuring they are trained on meaningful, relevant information — garbage in, garbage out.

Additionally, we can now see that DNAH6 appears only in the early-stage group. This suggests that DNAH6 may be a particularly valuable feature for helping the model distinguish between early and late cancer stages.


```{r}
# clinical_stage should already have: PATIENT_ID, Stage_num, Stage_grp, Stage_binary_2, etc.
# If not, this rebuilds it quickly:

clean_stage_to_14 <- function(x) {
  x <- toupper(as.character(x))
  x <- gsub("^STAGE\\s*", "", x)
  x <- gsub("[A]+$", "", x)
  x <- trimws(x)

  dplyr::case_when(
    x %in% c("I", "1")   ~ 1L,
    x %in% c("II", "2")  ~ 2L,
    x %in% c("III", "3") ~ 3L,
    x %in% c("IV", "4")  ~ 4L,
    TRUE ~ NA_integer_
  )
}

clinical_stage <- clinical %>%
  mutate(
    Stage_raw = AJCC_PATHOLOGIC_TUMOR_STAGE,
    Stage_num = clean_stage_to_14(Stage_raw),
    Stage_lowhigh = case_when(
      Stage_num %in% c(1, 2) ~ "Low Stage",   # Stage I–II
      Stage_num %in% c(3, 4) ~ "High Stage",  # Stage III–IV
      TRUE ~ NA_character_
    )
  ) %>%
  filter(!is.na(Stage_lowhigh)) %>%
  select(PATIENT_ID, Stage_num, Stage_lowhigh)



```










```{r}
library(dplyr)
library(e1071)

# --------------------------------
# 1. Gene list for the model
# --------------------------------

genes_for_model <- gene_diff_tbl$Gene[1:30]


# keep only genes that exist in mut_matrix
genes_for_model <- intersect(genes_for_model, rownames(mut_matrix))
genes_for_model

# --------------------------------
# 2. Build patient × gene + stage data
# --------------------------------

mut_df <- t(mut_matrix[genes_for_model, , drop = FALSE]) %>%
  as.data.frame() %>%
  mutate(PATIENT_ID = rownames(.))

# clinical_stage must have PATIENT_ID and Stage_lowhigh ("Low Stage", "High Stage")
df <- mut_df %>%
  inner_join(
    clinical_stage %>% select(PATIENT_ID, Stage_lowhigh),
    by = "PATIENT_ID"
  ) %>%
  filter(!is.na(Stage_lowhigh))

df$Stage_lowhigh <- factor(df$Stage_lowhigh)
table(df$Stage_lowhigh)

# --------------------------------
# 3. Stratified 70/30 train–test split
# --------------------------------

set.seed(13)

low  <- df %>% filter(Stage_lowhigh == "Low Stage")
high <- df %>% filter(Stage_lowhigh == "High Stage")

low_idx  <- sample(seq_len(nrow(low)),  size = floor(0.75 * nrow(low)))
high_idx <- sample(seq_len(nrow(high)), size = floor(0.75 * nrow(high)))

train <- bind_rows(low[low_idx, ],  high[high_idx, ])
test  <- bind_rows(low[-low_idx, ], high[-high_idx, ])

table(train$Stage_lowhigh)
table(test$Stage_lowhigh)

# --------------------------------
# 4. Fit Naive Bayes model
# --------------------------------

train_nb <- train[, c(genes_for_model, "Stage_lowhigh")]

nb_model <- naiveBayes(
  Stage_lowhigh ~ .,
  data = train_nb
)

# --------------------------------
# 5. Predict with probability threshold
# --------------------------------

# raw class probabilities on test set
nb_prob <- predict(nb_model, newdata = test[, genes_for_model, drop = FALSE],
                   type = "raw")

# choose threshold for "High Stage"
threshold <- 0.999 # <-- change this to 0.5, 0.7, etc. to tune behaviour

nb_pred <- ifelse(nb_prob[, "High Stage"] > threshold,
                  "High Stage", "Low Stage")

nb_pred <- factor(nb_pred, levels = levels(df$Stage_lowhigh))

# --------------------------------
# 6. Accuracy + confusion matrix
# --------------------------------

nb_accuracy <- mean(nb_pred == test$Stage_lowhigh)
nb_accuracy

nb_confusion <- table(Predicted = nb_pred, Actual = test$Stage_lowhigh)
nb_confusion

```


Because Naive Bayes tends to produce extremely high or low probabilities on this sparse mutation data, a very high decision threshold (0.999) was used for calling a tumor “High Stage”. This biases the classifier toward catching almost all truly high-stage tumors (sensitivity ≈ 94%), but at the cost of low specificity (≈ 32%) and modest overall accuracy (≈ 46%).


Could be a valid model because it rarely misses the high 

```{r}
library(pROC)
library(ggplot2)
library(dplyr)

# --------------------------------
# 1. Basic Metrics
# --------------------------------

nb_accuracy
nb_confusion

# Compute sensitivity, specificity, precision, recall, F1
TP <- nb_confusion["High Stage", "High Stage"]
TN <- nb_confusion["Low Stage",  "Low Stage"]
FP <- nb_confusion["High Stage", "Low Stage"]
FN <- nb_confusion["Low Stage",  "High Stage"]

sensitivity <- TP / (TP + FN)
specificity <- TN / (TN + FP)
precision   <- TP / (TP + FP)
recall      <- sensitivity
F1          <- 2 * (precision * recall) / (precision + recall)

metrics <- tibble(
  Metric = c("Accuracy", "Sensitivity (Recall)", "Specificity",
             "Precision", "F1 Score"),
  Value = c(nb_accuracy, sensitivity, specificity, precision, F1)
)

metrics

```







```{r}

library(dplyr)
library(e1071)

# --------------------------------
# 1. Gene list for the model
# --------------------------------

gene_freq <- rowSums(mut_matrix)
genes_for_model_2 <- names(sort(gene_freq, decreasing = TRUE))[1:30] # look at top 50 mutated genes


gene_freq <- rowSums(mut_matrix)
genes_for_model_2 <- names(sort(gene_freq, decreasing = TRUE))[1]


# keep only genes that exist in mut_matrix
genes_for_model_2 <- intersect(genes_for_model_2, rownames(mut_matrix))
genes_for_model_2

# --------------------------------
# 2. Build patient × gene + stage data
# --------------------------------

mut_df <- t(mut_matrix[genes_for_model_2, , drop = FALSE]) %>%
  as.data.frame() %>%
  mutate(PATIENT_ID = rownames(.))

genes_for_model

# clinical_stage must have PATIENT_ID and Stage_lowhigh ("Low Stage", "High Stage")
df <- mut_df %>%
  inner_join(
    clinical_stage %>% select(PATIENT_ID, Stage_lowhigh),
    by = "PATIENT_ID"
  ) %>%
  filter(!is.na(Stage_lowhigh))

df$Stage_lowhigh <- factor(df$Stage_lowhigh)
table(df$Stage_lowhigh)

# --------------------------------
# 3. Stratified 70/30 train–test split
# --------------------------------

set.seed(13)

low  <- df %>% filter(Stage_lowhigh == "Low Stage")
high <- df %>% filter(Stage_lowhigh == "High Stage")

low_idx  <- sample(seq_len(nrow(low)),  size = floor(0.75 * nrow(low)))
high_idx <- sample(seq_len(nrow(high)), size = floor(0.75 * nrow(high)))

train <- bind_rows(low[low_idx, ],  high[high_idx, ])
test  <- bind_rows(low[-low_idx, ], high[-high_idx, ])

table(train$Stage_lowhigh)
table(test$Stage_lowhigh)

# --------------------------------
# 4. Fit Naive Bayes model
# --------------------------------

train_nb <- train[, c(genes_for_model_2, "Stage_lowhigh")]

nb_model <- naiveBayes(
  Stage_lowhigh ~ .,
  data = train_nb
)

# --------------------------------
# 5. Predict with probability threshold
# --------------------------------

# raw class probabilities on test set
nb_prob <- predict(nb_model, newdata = test[, genes_for_model_2, drop = FALSE],
                   type = "raw")

# choose threshold for "High Stage"
threshold <- 0.999 # <-- change this to 0.5, 0.7, etc. to tune behaviour

nb_pred <- ifelse(nb_prob[, "High Stage"] > threshold,
                  "High Stage", "Low Stage")

nb_pred <- factor(nb_pred, levels = levels(df$Stage_lowhigh))

# --------------------------------
# 6. Accuracy + confusion matrix
# --------------------------------

nb_accuracy <- mean(nb_pred == test$Stage_lowhigh)
nb_accuracy

nb_confusion <- table(Predicted = nb_pred, Actual = test$Stage_lowhigh)
nb_confusion

```




```{r}
library(dplyr)
library(e1071)

# df must already exist and have:
# - one column per gene in genes_for_model
# - PATIENT_ID
# - Stage_lowhigh (factor with "Low Stage", "High Stage")

# ---- function: run Naive Bayes once for a given seed ----
run_nb_once <- function(seed, df, genes_for_model, threshold = 0.999) {
  set.seed(seed)
  
  # stratified split 75/25
  low  <- df %>% filter(Stage_lowhigh == "Low Stage")
  high <- df %>% filter(Stage_lowhigh == "High Stage")
  
  low_idx  <- sample(seq_len(nrow(low)),  size = floor(0.75 * nrow(low)))
  high_idx <- sample(seq_len(nrow(high)), size = floor(0.75 * nrow(high)))
  
  train <- bind_rows(low[low_idx, ],  high[high_idx, ])
  test  <- bind_rows(low[-low_idx, ], high[-high_idx, ])
  
  # fit NB on selected genes
  train_nb <- train[, c(genes_for_model, "Stage_lowhigh")]
  nb_model <- naiveBayes(Stage_lowhigh ~ ., data = train_nb)
  
  # predict probabilities
  nb_prob <- predict(
    nb_model,
    newdata = test[, genes_for_model, drop = FALSE],
    type = "raw"
  )
  
  # hard classification with threshold on P(High Stage)
  nb_pred <- ifelse(nb_prob[, "High Stage"] > threshold,
                    "High Stage", "Low Stage")
  nb_pred <- factor(nb_pred, levels = levels(df$Stage_lowhigh))
  
  # confusion matrix
  cm <- table(Predicted = nb_pred, Actual = test$Stage_lowhigh)
  
  # make sure all cells exist
  TP <- cm["High Stage", "High Stage"] %||% 0
  TN <- cm["Low Stage",  "Low Stage"]  %||% 0
  FP <- cm["High Stage", "Low Stage"]  %||% 0
  FN <- cm["Low Stage",  "High Stage"] %||% 0
  
  accuracy    <- (TP + TN) / (TP + TN + FP + FN)
  sensitivity <- if ((TP + FN) > 0) TP / (TP + FN) else NA
  specificity <- if ((TN + FP) > 0) TN / (TN + FP) else NA
  precision   <- if ((TP + FP) > 0) TP / (TP + FP) else NA
  F1          <- if (!is.na(precision) && !is.na(sensitivity) &&
                     (precision + sensitivity) > 0) {
    2 * precision * sensitivity / (precision + sensitivity)
  } else {
    NA
  }
  
  tibble(
    seed        = seed,
    accuracy    = accuracy,
    sensitivity = sensitivity,
    specificity = specificity,
    precision   = precision,
    F1          = F1
  )
}

# little helper for missing cells in confusion matrix
`%||%` <- function(x, y) if (is.null(x) || length(x) == 0) y else x

# ---- run over many seeds ----
seeds <- 1:50   # try 50 different random splits

nb_results <- purrr::map_dfr(
  seeds,
  ~run_nb_once(.x, df = df, genes_for_model = genes_for_model, threshold = 0.999)
)

# look at average + spread of metrics
summary_stats <- nb_results %>%
  summarise(
    across(
      c(accuracy, sensitivity, specificity, precision, F1),
      list(mean = mean, sd = sd),
      na.rm = TRUE
    )
  )

nb_results
summary_stats

```







