---
title: "Sarah"
author: "Sarah Dumont"
date: "2025-11-25"
output: html_document
---


```{r setup, message=FALSE, warning=FALSE}
knitr::opts_chunk$set(echo = TRUE)

# --- Setup ---
knitr::opts_chunk$set(echo = TRUE)
library(dplyr)
library(knitr)
library(readr)
library(tibble)
library(ggplot2)

# --- Load Data ---
clinical <- read_tsv(
  "data_clinical_patient.txt",
  comment = "#",
  show_col_types = FALSE
)

mutation <- read.delim(
  "data_mutations.txt",
  sep = "\t",
  header = TRUE
)



# --- Helper Function ---
normalize_tcga <- function(x) {
  x <- as.character(x)
  x <- gsub("\\.", "-", x)
  substr(x, 1, 12)
}

# --- Unique IDs ---
clin_ids <- unique(normalize_tcga(clinical$PATIENT_ID))
mut_ids <- unique(normalize_tcga(mutation$Tumor_Sample_Barcode))
rna_cols <- setdiff(colnames(rnaseq), c("Hugo_Symbol", "Entrez_Gene_Id"))
rna_ids <- unique(normalize_tcga(rna_cols))

# --- Count Table ---
counts_tbl <- tibble(
  Dataset = c("Clinical", "Mutation", "RNAseq"),
  `Unique Patients` = c(length(clin_ids), length(mut_ids), length(rna_ids))
)

kable(
  counts_tbl,
  align = c("l", "c"),
  caption = "Unique Patient Counts per Dataset"
)

# --- Patients in all three ---
common_ids <- Reduce(intersect, list(clin_ids, mut_ids, rna_ids))

kable(
  tibble(`Patients with ALL Clinical + Mutation + RNAseq` = length(common_ids)),
  align = "c",
  caption = "Patients with All Three Data Types"
)

# ----------------------------
# Filter clinical to common patients
# ----------------------------

clinical <- clinical %>%
  mutate(PATIENT_ID = normalize_tcga(PATIENT_ID)) %>%
  filter(PATIENT_ID %in% common_ids)

# ----------------------------
# Non-synonymous mutation filtering
# ----------------------------

nonsyn_types <- c(
  "Missense_Mutation", "Nonsense_Mutation", "Nonstop_Mutation",
  "Frame_Shift_Del", "Frame_Shift_Ins",
  "In_Frame_Del", "In_Frame_Ins",
  "Splice_Site", "Translation_Start_Site"
)

mutation <- mutation %>%
  mutate(PATIENT_ID = normalize_tcga(Tumor_Sample_Barcode)) %>%
  filter(
    Variant_Classification %in% nonsyn_types,
    PATIENT_ID %in% common_ids
  )


```



```{r}
colnames(clinical)
```




```{r}



### 🔹 Chunk 3 — Filter for *Non-Synonymous* Mutations (your specified list)

nonsynonymous <- c(
  "Missense_Mutation", "Nonsense_Mutation",
  "Frame_Shift_Del", "Frame_Shift_Ins",
  "In_Frame_Del", "In_Frame_Ins",
  "Splice_Site", "Splice_Region", "Nonstop_Mutation", 
  "Translation_Start_Site"
)

mut_filtered <- mutations[mutations$Variant_Classification %in% nonsynonymous, ]

```



```{r}


### 🔹 Chunk 4 — Build Gene × Patient Matrix + TMB

genes    <- sort(unique(mut_filtered$Hugo_Symbol))
patients <- sort(unique(mut_filtered$Patient_ID))

# Matrix: rows = genes, cols = patients, entries = 0/1 for "has non-syn mutation"
mut_matrix <- matrix(
  0L,
  nrow = length(genes),
  ncol = length(patients),
  dimnames = list(genes, patients)
)

for (i in seq_len(nrow(mut_filtered))) {
  g <- mut_filtered$Hugo_Symbol[i]
  p <- mut_filtered$Patient_ID[i]
  mut_matrix[g, p] <- 1L
}

# Tumor Mutation Burden (TMB) = number of mutated genes per patient
TMB <- colSums(mut_matrix)

tmb_df <- data.frame(
  PATIENT_ID = names(TMB),
  TMB = as.numeric(TMB),
  row.names = NULL
)

```





```{r}
# --- Find and collapse cancer stage into 1–4 ---
stage_col <- "AJCC_PATHOLOGIC_TUMOR_STAGE"

stage_col <- stage_candidates[1]     # use first match
stage_col                             # prints the chosen column

# 2) Function to clean stage strings
clean_stage_to_14 <- function(x) {
  x <- toupper(as.character(x))
  x <- gsub("^STAGE\\s*", "", x)   # remove "Stage "
  x <- gsub("[A]+$", "", x)      # remove letters A/B/C/etc.
  x <- trimws(x)

  dplyr::case_when(
    x %in% c("I", "1")   ~ 1L,
    x %in% c("II", "2")  ~ 2L,
    x %in% c("III", "3") ~ 3L,
    x %in% c("IV", "4")  ~ 4L,
    TRUE ~ NA_integer_
  )
}

# 3) Apply stage collapse
clinical_stage <- clinical %>%
  mutate(
    Stage_raw = .data[[stage_col]],
    Stage_num = clean_stage_to_14(Stage_raw),
    Stage_grp = factor(
      Stage_num,
      levels = 1:4,
      labels = paste("Stage", 1:4)
    )
  ) %>%
  filter(!is.na(Stage_num)) %>%
  select(PATIENT_ID, Stage_raw, Stage_num, Stage_grp)

clinical_stage


```


```{r}


tmb_stage <- tmb_df %>%
  inner_join(clinical_stage, by = "PATIENT_ID")

ggplot(tmb_stage, aes(x = Stage_grp, y = TMB)) +
  geom_boxplot(fill = "lightblue", alpha = 0.5) +
  geom_jitter(width = 0.2, alpha = 0.6) +
  theme_minimal(base_size = 14) +
  labs(
    title = "Tumor Mutation Burden by Collapsed Stage (I–IV)",
    x = "Stage (Grouped)",
    y = "TMB (Non-synonymous Mutations)"
  )

```


```{r}
# Count patients per collapsed stage
table(clinical_stage$Stage_grp)

# Or a nicer data frame
stage_counts <- clinical_stage %>%
  count(Stage_grp)

stage_counts

```
Troubles: cannot do with by simply splitting stages because stage for only has 4 patients so we will combine - look at stage 1, stage 2 and late stages 3 + 4

```{r}
# Collapse Stage into Stage 1, Stage 2, Late Stages (3+4)
clinical_stage <- clinical_stage %>%
  mutate(
    Stage_binary = case_when(
      Stage_num == 1 ~ "Stage 1",
      Stage_num == 2 ~ "Stage 2",
      Stage_num %in% c(3, 4) ~ "Late Stages",
      TRUE ~ NA_character_
    ),
    Stage_binary = factor(
      Stage_binary,
      levels = c("Stage 1", "Stage 2", "Late Stages")
    )
  )

```

```{r}
# Merge TMB with early/late stage
tmb_stage_binary <- tmb_df %>%
  inner_join(clinical_stage, by = "PATIENT_ID") %>%
  filter(!is.na(Stage_binary))

# Plot
ggplot(tmb_stage_binary, aes(x = Stage_binary, y = TMB)) +
  geom_boxplot(fill = "lightblue", alpha = 0.6) +
  geom_jitter(width = 0.2, alpha = 0.5) +
  theme_minimal(base_size = 14) +
  labs(
    title = "Tumor Mutation Burden: Early vs Late Stage",
    x = "Stage Group",
    y = "TMB (Non-synonymous Mutations)"
  )


```
TMB alone is not a good staging biomarker
These box plots do not have enough variation a model will not be able to distiguish these catigoeies   


```{r}
gene_freq <- rowSums(mut_matrix)
top_genes <- names(sort(gene_freq, decreasing = TRUE))[1:20]
top_genes

# Build patient × gene table
gene_df <- t(mut_matrix[top_genes, ]) %>%
  as.data.frame() %>%
  mutate(PATIENT_ID = rownames(.)) %>%
  inner_join(clinical_stage, by = "PATIENT_ID")

# Calculate mutation proportion per stage
stage_gene_freq <- gene_df %>%
  group_by(Stage_binary) %>%
  summarise(across(top_genes, mean))

stage_gene_freq
write.table(stage_gene_freq, row.names = FALSE, sep = ",")



```
Now we can see there is diffrences between these but if there enough to create a modle that is going to have high accuray where high is defiend as above 90 percent. Short answer no. There is not enough spread in the data. Lets try something else lets look just at early and late stages




```{r}

# Collapse Stage into Stage 1, Stage 2, Late Stages (3+4)
clinical_stage <- clinical_stage %>%
  mutate(
    Stage_binary_2 = case_when(
      Stage_num %in% c(1, 2) ~ "Early Stages",
      Stage_num %in% c(3, 4) ~ "Late Stages",
      TRUE ~ NA_character_
    ),
    Stage_binary_2 = factor(
      Stage_binary_2,
      levels = c("Early Stages", "Late Stages")
    )
  )

```


```{r}
gene_freq <- rowSums(mut_matrix)
top_genes <- names(sort(gene_freq, decreasing = TRUE))[1:50]
top_genes

# Build patient × gene table
gene_df <- t(mut_matrix[top_genes, ]) %>%
  as.data.frame() %>%
  mutate(PATIENT_ID = rownames(.)) %>%
  inner_join(clinical_stage, by = "PATIENT_ID")

# Calculate mutation proportion per stage
stage_gene_freq <- gene_df %>%
  group_by(Stage_binary_2) %>%
  summarise(across(top_genes, mean))

stage_gene_freq
#write.table(stage_gene_freq, row.names = FALSE, sep = ",")
```


```{r}

# Get just the gene columns (everything except Stage_binary_2)
genes <- setdiff(names(stage_gene_freq), "Stage_binary_2")

# Extract rows for Early and Late
early_row <- stage_gene_freq %>%
  filter(Stage_binary_2 == "Early Stages") %>%
  select(all_of(genes))

late_row <- stage_gene_freq %>%
  filter(Stage_binary_2 == "Late Stages") %>%
  select(all_of(genes))

# Compute Late - Early for each gene
diff_vec <- as.numeric(late_row[1, ] - early_row[1, ])

# Build a table: gene, difference, absolute difference
gene_diff_tbl <- tibble(
  Gene     = genes,
  Diff     = diff_vec,           # Late - Early (signed)
  Abs_Diff = abs(diff_vec)       # absolute difference
) %>%
  arrange(desc(Abs_Diff))         # sort by largest absolute difference

gene_diff_tbl <- gene_diff_tbl[1:10, ]

# If you want an easy copy/paste version:
write.table(gene_diff_tbl, row.names = FALSE, sep = ",")

```
```{r}
# Plot mutation proportion for Top 10 different genes
top10_genes <- gene_diff_tbl$Gene[1:4]

bar_df <- stage_gene_freq %>%
  select(Stage_binary_2, all_of(top10_genes)) %>%
  pivot_longer(cols = all_of(top10_genes),
               names_to = "Gene",
               values_to = "Proportion")

ggplot(bar_df, aes(x = Gene, y = Proportion, fill = Stage_binary_2)) +
  geom_col(position = "dodge") +
  theme_minimal(base_size = 14) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  labs(
    title = "Mutation Proportions of Top 4 Differentiating Genes",
    y = "Proportion Mutated"
  )

```


This shows that after performing some preliminary re-organization of the data, we now have a structured set of mutation features that can more effectively separate cancer stages into early and late groups. This improves the usefulness of the dataset and should increase the accuracy of any future predictive models by ensuring they are trained on meaningful, relevant information — garbage in, garbage out.

Additionally, we can now see that DNAH6 appears only in the early-stage group. This suggests that DNAH6 may be a particularly valuable feature for helping the model distinguish between early and late cancer stages.




```{r}
colnames(clinical)

```

  

```{r}
unique_counts <- sapply(clinical, function(x) length(unique(na.omit(x))))
unique_counts

```

```{r}
clinical$PATH_T_STAGE
clinical$PATH_N_STAGE
clinical$PATH_M_STAGE


```
```{r}
library(dplyr)
library(stringr)

# ---- CLEAN T STAGE ----
clean_T <- function(x) {
  x <- toupper(as.character(x))
  x <- gsub("[A-Z]+$", "", x)     # remove A/B/C, etc.
  x <- gsub("^T", "", x)          # remove leading T
  x <- trimws(x)

  case_when(
    x %in% c("0","1","2","3","4") ~ paste0("T", x),
    TRUE ~ NA_character_
  )
}

clinical$T_clean <- clean_T(clinical$PATH_T_STAGE)

T_counts <- table(clinical$T_clean, useNA = "ifany")
T_counts


# ---- CLEAN N STAGE ----
clean_N <- function(x) {
  x <- toupper(as.character(x))
  x <- gsub("^N", "", x)          # remove leading N
  x <- trimws(x)

  case_when(
    x %in% c("0","1", "X") ~ paste0("N", x),
    TRUE ~ NA_character_
  )
}

clinical$N_clean <- clean_N(clinical$PATH_N_STAGE)

N_counts <- table(clinical$N_clean, useNA = "ifany")
N_counts


# ---- CLEAN M STAGE ----
clean_M <- function(x) {
  x <- toupper(as.character(x))
  x <- gsub("^M", "", x)          # remove leading M
  x <- trimws(x)

  case_when(
    x %in% c("0","1", "X") ~ paste0("M", x),
    TRUE ~ NA_character_
  )
}

clinical$M_clean <- clean_M(clinical$PATH_M_STAGE)

M_counts <- table(clinical$M_clean, useNA = "ifany")
M_counts


# ---- FILTER OUT NA T/N/M IF NEEDED ----
clinical_TNM <- clinical %>%
  filter(!is.na(T_clean), !is.na(N_clean), !is.na(M_clean))

```
so from this it is clear that T0 and T1 need to be combined as well as T3 and T4
and N1 and M1 need to be ommited and cannot be correctly prodicited with this data set. Upon further reserach anything with the compenet X so NX and MX means that this data was not collected meaning There so not enough data to do analysis on for these two categories 


```{r}

clinical <- clinical %>%
  mutate(
    T_clean = clean_T(PATH_T_STAGE),
  )


# --- APPLY YOUR DECISIONS BASED ON COUNTS ---
# - Combine T0 + T1
# - Combine T3 + T4
# - Drop N1 (too few)
# - Drop M1 (too few)


clinical_TNM_collapsed <- clinical %>%
  mutate(
    T_group = case_when(
      T_clean %in% c("T0", "T1")      ~ "T0-1",
      T_clean == "T2"                 ~ "T2",
      T_clean %in% c("T3", "T4")      ~ "T3-4",
      TRUE                            ~ NA_character_
    )
  ) %>%
  # drop unusable N and M categories
  filter(
    !is.na(T_group),
    N_clean != "N1",
    M_clean != "M1",

  )

# Check final distributions
table(clinical_TNM_collapsed$T_group)
table(clinical_TNM_collapsed$N_clean)
table(clinical_TNM_collapsed$M_clean)
  
```

```{r}
library(dplyr)

## 1. Create collapsed T groups: T0-1, T2, T3-4 ----

clean_T <- function(x) {
  x <- toupper(as.character(x))
  x <- gsub("[A-Z]+$", "", x)   # drop suffix A/B/C
  x <- gsub("^T", "", x)        # drop leading T
  x <- trimws(x)

  dplyr::case_when(
    x %in% c("0", "1") ~ "T0-1",
    x == "2"           ~ "T2",
    x %in% c("3", "4") ~ "T3-4",
    TRUE               ~ NA_character_
  )
}

clinical <- clinical %>%
  mutate(T_group = clean_T(PATH_T_STAGE))

## 2. Keep only patients with valid T_group and present in mut_matrix ----

clinical_t <- clinical %>%
  filter(!is.na(T_group))

common_ids <- intersect(colnames(mut_matrix), clinical_t$PATIENT_ID)

mut_t <- mut_matrix[, common_ids, drop = FALSE]

clinical_t <- clinical_t %>%
  filter(PATIENT_ID %in% common_ids) %>%
  arrange(match(PATIENT_ID, colnames(mut_t)))

## 3. Remove genes with zero variance ----

gene_var <- apply(mut_t, 1, var)
mut_t_var <- mut_t[gene_var > 0, , drop = FALSE]

## 4. PCA on mutation data only (rows = patients, cols = genes) ----

X <- t(mut_t_var)  # patients × genes

pca_res <- prcomp(X, center = TRUE, scale. = TRUE)

## 5. Build PCA data frame with *all* PCs + T_group ----
##    (number of PCs = nrow(X) - 1)

all_pcs <- ncol(pca_res$x)

pca_df <- data.frame(
  pca_res$x[, 1:all_pcs, drop = FALSE],
  T_group = clinical_t$T_group
)



```



```{r}

# Basic 2D PCA plot: PC1 vs PC2 colored by T_group
ggplot(pca_df, aes(x = PC1, y = PC2, color = T_group)) +
  geom_point(alpha = 0.8, size = 2) +
  theme_minimal(base_size = 14) +
  labs(
    title = "PCA of Mutation Data (PC1 vs PC2)",
    x = "PC1",
    y = "PC2"
  )

# PC2 vs PC3 (sometimes separation is clearer on other components)
ggplot(pca_df, aes(x = PC2, y = PC3, color = T_group)) +
  geom_point(alpha = 0.8, size = 2) +
  theme_minimal(base_size = 14) +
  labs(
    title = "PCA of Mutation Data (PC2 vs PC3)",
    x = "PC2",
    y = "PC3"
  )

# Optional: facet by T_group to see each cluster separately on PC1/PC2
ggplot(pca_df, aes(x = PC1, y = PC2)) +
  geom_point(alpha = 0.8, size = 2) +
  facet_wrap(~ T_group) +
  theme_minimal(base_size = 14) +
  labs(
    title = "PCA (PC1 vs PC2) Faceted by T Group",
    x = "PC1",
    y = "PC2"
  )

```

```{r}
# Install (if needed)
install.packages("umap")

library(umap)
library(ggplot2)
library(dplyr)

# --- UMAP input matrix ---
# X = patients × genes (same as PCA input)
X <- t(mut_t_var)   # rows = patients, cols = genes

# --- Run UMAP ---
set.seed(123)
umap_res <- umap(X)

# Build UMAP DF
umap_df <- data.frame(
  UMAP1 = umap_res$layout[,1],
  UMAP2 = umap_res$layout[,2],
  T_group = clinical_t$T_group
)

# --- Plot UMAP ---
ggplot(umap_df, aes(x = UMAP1, y = UMAP2, color = T_group)) +
  geom_point(alpha = 0.8, size = 2) +
  theme_minimal(base_size = 14) +
  labs(
    title = "UMAP of Mutation Data (Colored by T Group)",
    x = "UMAP1",
    y = "UMAP2"
  )

```

```{r}
library(nnet)

# Use PC1–PC10 as predictors
model_data <- pca_df
model_data$T_group <- factor(model_data$T_group)

set.seed(123)
logit_model <- multinom(T_group ~ ., data = model_data)

summary(logit_model)

# Predictions
pred <- predict(logit_model, model_data)

# Accuracy
mean(pred == model_data$T_group)

```
```{r}
install.packages("randomForest")
library(randomForest)


```

```{r}
# Variance explained by each PC
var_explained <- pca_res$sdev^2 / sum(pca_res$sdev^2)

# Cumulative variance
cum_var <- cumsum(var_explained)

# Find number of PCs needed for 90% variance
which(cum_var >= 0.20)[1]

```

```{r}


k <- 20  # <- change this to use 5, 10, 20, ... PCs

model_data <- pca_df %>%
  select(all_of(paste0("PC", 1:k)), T_group)
model_data$T_group <- factor(model_data$T_group)

set.seed(123)

n <- nrow(model_data)
idx <- sample(seq_len(n), size = floor(0.7 * n))   # 70% training

train <- model_data[idx, ]
test  <- model_data[-idx, ]

# ------------------------
# Random Forest Model
# ------------------------
rf_model <- randomForest(
  T_group ~ ., 
  data = train,
  ntree = 500,       # number of trees
  mtry = 3,          # PCs to try at each split (tuneable)
  importance = TRUE
)

# ------------------------
# Predictions + Accuracy
# ------------------------
rf_pred <- predict(rf_model, test)

rf_accuracy <- mean(rf_pred == test$T_group)
rf_accuracy


```
```{r}
#library(caret)
rf_confusion <- table(Predicted = rf_pred, Actual = test$T_group)
rf_confusion
```

